SCPCL000822 - Validation of tumor cells

### 2024-06-12 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
break out heatmap functions and explore the unknown

@@

# Set default ggplot theme
theme_set(
  theme_bw()
)
```

@@

# pull out the UMAP coordinates and make a data frame to use for plotting
coldata_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = "ensembl_gene_id") |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
  # add in marker gene classifications
  dplyr::left_join(marker_gene_results_df, by = "barcodes") |> 
  # add in copykat results 
  dplyr::left_join(copykat_results_df, by = "barcodes") |> 
  # add in infercnv results and metadata
  dplyr::left_join(infercnv_results_df, by = "barcodes") |> 
  dplyr::left_join(infercnv_metadata_df, by = "barcodes") |> 
  # gene set scores 
  dplyr::left_join(geneset_scores_df, by = "barcodes")

@@

```{r}
# plot tumor cells for each classification method 
tumor_umap_df <- coldata_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, marker_gene_classification, copykat_classification, cnv_proportion_classification) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  )

@@

This plot shows us that there is a pretty clear consensus that most of the cells in the bottom left group are tumor cells. 
From the original marker gene analysis we know that not all marker genes are expressed and that there is quite a bit of variance in expression. 
Because of that and the fact that both CNV inference methods show similar annotations, we will classify any cells that are labeled as tumor by both `CopyKAT` and `InferCNV` as tumor cells.

```{r}
# create a combined classification between inferCNV and copyKAT
# cells that are classified as tumor by both are kept as tumor, all other are normal 
coldata_df <- coldata_df |> 
  dplyr::mutate(
    combined_cnv_classification = dplyr::if_else(
      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor",
      "Tumor",
      "Normal"
    ),
    # there are cells classified as NA by copykat, we want those to be normal 
    combined_cnv_classification = dplyr::if_else(is.na(copykat_classification) & is.na(combined_cnv_classification), "Normal", combined_cnv_classification)
  )
```

```{r}
# look at which cells are tumor/normal
# be sure to remove any duplicates 
coldata_df |> 
  dplyr::select(UMAP1, UMAP2, barcodes, combined_cnv_classification) |>
  dplyr::distinct() |> 
  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

The remainder of the notebook will focus on validating that these in fact are tumor cells and all other cells are classified as normal.

@@

```{r}
# calculate sum of the scaled gene expression values for plotting 
marker_density_df <- coldata_df |> 
  dplyr::select(barcodes, gene_symbol, gene_expression, ends_with("classification")) |> 
  unique() |> 
  dplyr::group_by(gene_symbol) |>
  # get z-scores for each gene
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |> 
  dplyr::ungroup() |> 
  dplyr::group_by(barcodes) |> 
  # add all the scores together 
  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |>
  # expand classifications column for easy faceting 
  dplyr::ungroup() |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  )

@@

@@

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- coldata_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

# create annotation df, make sure that dimensions match up by only getting unique barcodes 
annotation_df <- coldata_df |> 
  dplyr::select(barcodes, combined_cnv_classification) |> 
  unique()

annotation <- ComplexHeatmap::columnAnnotation(
  classification = annotation_df$combined_cnv_classification,
  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05")))
```

```{r}
# plot heatmap of marker genes 
ComplexHeatmap::Heatmap(
  marker_gene_heatmap,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "Marker gene symbol",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  row_names_side = "left",
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = TRUE,
  show_column_names = FALSE,
  bottom_annotation = annotation
)
```

As expected, the tumor cells appear to have more marker genes that are expressed than normal cells.

@@

```{r}
# plot tumor cells for each classification method 
geneset_plot_df <- coldata_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  )

@@

@@

ComplexHeatmap::Heatmap(
  geneset_heatmap,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  row_names_side = "left",
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = TRUE,
  show_column_names = FALSE,
  bottom_annotation = annotation
)
```

@@

@@

```{r}
# get matrix of barcodes by chromosome
gain_cnv_df <- coldata_df |> 
  dplyr::select(barcodes, starts_with("proportion_scaled_dupli")) |> 
  unique() |> 
  tibble::column_to_rownames("barcodes") |> 
  as.matrix()

# remove extra characters in colnames, proportion_scaled_dupli
colnames(gain_cnv_df) <- stringr::word(colnames(gain_cnv_df), -1, sep = "_")

# need to make a new annotation, since we want to annotate the rows here 
annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05")))
```

```{r}
ComplexHeatmap::Heatmap(
  gain_cnv_df,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  show_row_names = FALSE,
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = FALSE,
  show_column_names = TRUE,
  right_annotation = annotation
)
```

@@

```{r}
# barcode by chromosome mtx with losses 
loss_cnv_df <- coldata_df |> 
  dplyr::select(barcodes, starts_with("proportion_scaled_loss")) |> 
  unique() |> 
  tibble::column_to_rownames("barcodes") |> 
  as.matrix()

colnames(loss_cnv_df) <- stringr::word(colnames(loss_cnv_df), -1, sep = "_")
```

```{r}
ComplexHeatmap::Heatmap(
  loss_cnv_df,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  show_row_names = FALSE,
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = FALSE,
  show_column_names = TRUE,
  right_annotation = annotation
)
```

@@

## Save final tumor annotations

@@

```{r}
# export final TSV with annotations 
final_classifications <- coldata_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    tumor_cell_classification = combined_cnv_classification
  ) |> 
  unique()

>>>>

## Define functions

```{r}
# create a heatmap where rows are marker genes or gene sets and columns are cells 
# this function adds a column annotation since cells are columns 
plot_gene_heatmap <- function(df, 
                              row_title = "",
                              legend_title = "", 
                              annotation = NULL){

  # plot heatmap of marker genes 
  heatmap <- ComplexHeatmap::Heatmap(
    df,
    # set the color scale based on min and max values
    col = circlize::colorRamp2(seq(min(df), max(df), length = 2), colors = c("white", "#00274C")),
    border = TRUE,
    ## Row parameters
    cluster_rows = TRUE,
    row_title = row_title,
    row_title_gp = grid::gpar(fontsize = 12),
    row_title_side = "left",
    row_names_side = "left",
    row_dend_side = "right",
    row_names_gp = grid::gpar(fontsize = 10),
    ## Column parameters
    cluster_columns = TRUE,
    show_column_names = FALSE,
    bottom_annotation = annotation,
    heatmap_legend_param = list(
      title = legend_title
    )
  )

  return(heatmap)
}

# create a heatmap where rows are cells and columns are chromosomes 
# this function adds a row annotation since cells are rows 
plot_cnv_heatmap <- function(df,
                             cnv_col, 
                             annotation = NULL,
                             legend_title = ""){

  # barcode by chromosome df 
  cnv_df <- df |> 
    dplyr::select(barcodes, starts_with(cnv_col)) |> 
    unique() |> 
    tibble::column_to_rownames("barcodes")

  # just get the name of the chromosome as column name 
  colnames(cnv_df) <- stringr::word(colnames(cnv_df), -1, sep = "_")

  # do a little bit of column reordering before converting to a matrix
  cnv_df <- cnv_df |> 
    dplyr::relocate(chr10, .before = chr11) |> 
    dplyr::relocate(chr19, .before = chr20) |> 
    dplyr::relocate(chr21, .before = chr22) |> 
    as.matrix()

  heatmap <- ComplexHeatmap::Heatmap(
    cnv_df,
    col = circlize::colorRamp2(seq(min(cnv_df), max(cnv_df), length = 2), colors = c("white", "#00274C")),
    border = TRUE,
    ## Row parameters
    cluster_rows = TRUE,
    row_title = "",
    row_title_gp = grid::gpar(fontsize = 12),
    row_title_side = "left",
    show_row_names = FALSE,
    row_dend_side = "right",
    row_names_gp = grid::gpar(fontsize = 10),
    ## Column parameters
    cluster_columns = FALSE,
    show_column_names = TRUE,
    right_annotation = annotation,
    heatmap_legend_param = list(
      title = legend_title
    )
  )
  return(heatmap)
}

@@

# Set default ggplot theme
theme_set(
  theme_bw()
)

# use jpegs for images 
knitr::opts_chunk$set(
  dev = "jpeg"
)

# quiet messages
options(readr.show_col_types = FALSE)
ComplexHeatmap::ht_opt(message = FALSE)
```

@@

# pull out the UMAP coordinates, add in marker genes and classifications 
# and make a data frame to use for plotting
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = "ensembl_gene_id") |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
  # add in marker gene classifications
  dplyr::left_join(marker_gene_results_df, by = "barcodes") |> 
  # add in copykat results 
  dplyr::left_join(copykat_results_df, by = "barcodes") |> 
  # add in infercnv results and metadata
  dplyr::left_join(infercnv_results_df, by = "barcodes") |> 
  dplyr::left_join(infercnv_metadata_df, by = "barcodes") |> 
  # gene set scores 
  dplyr::left_join(geneset_scores_df, by = "barcodes")

@@

- Marker genes only: For each cell, the expression of all marker genes was transformed using a z-transformation and summed together. 
Any cells with a total marker gene expression value > 0 were called as tumor cells. 
- CopyKAT - no reference: [`CopyKAT`](https://github.com/navinlabcode/copykat) was run without indicating any reference cells and cells were classified as aneuploid (Tumor) or diploid (Normal). 
- InferCNV - [`InferCNV`](https://github.com/broadinstitute/inferCNV/wiki) was used to call copy number variations using endothelial and fibroblasts assigned by `CellAssign` as the normal reference group. 
This returns a proportion of each chromosome with a CNV detected.
We then calculate the genomic CNV proportion for each cell across all chromosomes, weighted by the number of genes in a chromosome.
Cells with a genomic CNV proportion greater than the mean  CNV proportion across all cells are called as tumor cells.

```{r}
# plot tumor cells for each classification method 
tumor_umap_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, marker_gene_classification, copykat_classification, cnv_proportion_classification) |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "copykat_classification" ~ "CopyKAT - no reference",
      method == "cnv_proportion_classification" ~ "InferCNV proportion"
    )
  )

@@

This plot shows us that there is a pretty clear consensus that most of the cells in the bottom right group are tumor cells. 
From the original marker gene analysis we know that not all marker genes are expressed and that there is quite a bit of variance in expression. 
Because of that and the fact that both CNV inference methods show similar annotations, we will classify any cells that are labeled as tumor by both `CopyKAT` and `InferCNV` as tumor cells.

```{r}
# create a combined classification between inferCNV and copyKAT
# cells that are classified as tumor by both are kept as tumor
# classified by both as normal are normal 
# in between are classified as unknown
classification_df <- classification_df |> 
  dplyr::mutate(
    combined_cnv_classification = dplyr::case_when(
      # only tumor if both call tumor 
      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor" ~ "Tumor",
      # only normal if both call normal and/or reference
      cnv_proportion_classification == "Normal" & copykat_classification == "Normal" ~ "Normal",
      cnv_proportion_classification == "Reference" & copykat_classification == "Normal" ~ "Normal",
      # if normal/tumor mismatch set to unknown 
      cnv_proportion_classification == "Normal" & copykat_classification == "Tumor" ~ "Unknown",
      cnv_proportion_classification == "Tumor" & copykat_classification == "Normal" ~ "Unknown",
      cnv_proportion_classification == "Reference" &  copykat_classification == "Tumor" ~ "Unknown", 
      !is.na(cnv_proportion_classification) & is.na(copykat_classification) ~ "Unknown", 
      is.na(cnv_proportion_classification) & !is.na(copykat_classification) ~ "Unknown",
      is.na(cnv_proportion_classification) & is.na(copykat_classification) ~ "Unknown"
    ))
```

```{r}
# look at which cells are tumor/normal
# be sure to remove any duplicates 
classification_df |> 
  dplyr::select(UMAP1, UMAP2, barcodes, combined_cnv_classification) |>
  dplyr::distinct() |> 
  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

The remainder of the notebook will focus on validating that these in fact are tumor cells and all other cells are classified as normal. 
We will also look at where the unknown cells fall in relation to tumor vs. normal.

@@

Just a note that the `marker_gene_classification` is a circular since these same marker genes were used to classify those cells.

```{r}
# calculate sum of the scaled gene expression values for plotting 
marker_density_df <- classification_df |> 
  dplyr::select(barcodes, gene_symbol, gene_expression, copykat_classification, cnv_proportion_classification, marker_gene_classification, combined_cnv_classification) |> 
  unique() |> 
  dplyr::group_by(gene_symbol) |>
  # get z-scores for each gene
  dplyr::mutate(transformed_gene_expression = scale(gene_expression)[, 1]) |> 
  dplyr::ungroup() |> 
  dplyr::group_by(barcodes) |> 
  # add all the scores together 
  dplyr::mutate(sum_exp = sum(transformed_gene_expression)) |>
  # expand classifications column for easy faceting 
  dplyr::ungroup() |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "copykat_classification" ~ "CopyKAT - no reference",
      method == "cnv_proportion_classification" ~ "InferCNV proportion",
      method == "combined_cnv_classification" ~ "CopyKAT and InferCNV"
    )
  )

@@

Another note is that the "Unknown" class appears to have marker gene expression more similar to the normal cells.

@@

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- classification_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

# create annotation df, make sure that dimensions match up by only getting unique barcodes 
annotation_df <- classification_df |> 
  dplyr::select(barcodes, combined_cnv_classification) |> 
  unique()

annotation <- ComplexHeatmap::columnAnnotation(
  classification = annotation_df$combined_cnv_classification,
  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
```

```{r}
# plot heatmap of marker genes 
plot_gene_heatmap(marker_gene_heatmap, 
                  row_title = "Marker gene symbol",
                  legend_title = "Marker gene \nexpression",
                  annotation = annotation)
```

As expected, the tumor cells appear to have more marker genes that are expressed than normal cells. 
The "Unknown" cells appear to have a small group of cells that cluster with the tumor cells on the right hand side of the plot.

@@

```{r}
# plot tumor cells for each classification method 
geneset_plot_df <- classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, ends_with("classification"), starts_with("mean")) |> 
  tidyr::pivot_longer(
    cols = starts_with("mean"),
    names_to = "geneset",
    values_to = "mean_score"
  ) |> 
  dplyr::mutate(
    geneset = stringr::word(geneset, -1, sep = "-")
  )

@@

We also see that the "Unknown" cells identified in the combined classification mostly mirror the gene set expression scores for normal cells with maybe a small group of cells having more similar gene set scores to tumor cells.

@@

# plot heatmap of gene set score 
plot_gene_heatmap(geneset_heatmap,
                  annotation = annotation,
                  legend_title = "Gene set \nscore")
```

@@

Again, most of the "Unknown" tumor cells tend to align with normal cells, with a few grouping in the area where cells transition from tumor to normal.

@@

```{r}
# need to make a new annotation, since we want to annotate the rows here 
annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))

# plot gains in cnv
plot_cnv_heatmap(classification_df,
                 cnv_col = "proportion_scaled_dupli",
                 annotation = annotation,
                 legend_title = "Copy number \ngain")
```

@@

```{r}
# plot loss in cnv
plot_cnv_heatmap(classification_df, 
                 cnv_col = "proportion_scaled_loss", 
                 annotation = annotation, 
                 legend_title = "Copy number \nloss")
```

@@

In both of these plots, the majority of "Unknown" cells cluster between tumor and normal or with the normal cells.

## Create final tumor annotations

In looking at the plots, it looks like there might be some of the "Unknown" cells that should actually be classified as tumor cells. 
In particular, we expect those that have a gain in Chr8 are tumor cells, so let's just look at Chr8 gains and see if we can find a cut off to use to categorize our "Unknown" cells.

```{r}
#density plot of chr8 gains
ggplot(classification_df, aes(x = proportion_scaled_dupli_chr8, fill = combined_cnv_classification)) +
  geom_density() +
  facet_grid(rows = vars(combined_cnv_classification))
```

It looks like we have a pretty clear distinction between the `proportion_scaled_dupli_chr8` for normal and tumor cells and we could call any cell with a value > 0.25 as tumor cells and those that are less than 0.25 as normal cells.

```{r}
new_classification_df <- classification_df |> 
  # remove duplicat
  dplyr::select(-c(gene_symbol, ensembl_gene_id, gene_expression)) |> 
  dplyr::distinct() |> 
  dplyr::mutate(
    combined_cnv_classification = dplyr::case_when(
      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 >= 0.25 ~ "Tumor",
      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 < 0.25 ~ "Normal",
      .default = combined_cnv_classification
    )
  )
```

```{r}
# account for new annotations 
annotation <- ComplexHeatmap::rowAnnotation(classification = new_classification_df$combined_cnv_classification,
                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))

# recreate the cnv heatmap 
plot_cnv_heatmap(new_classification_df,
                 cnv_col = "proportion_scaled_dupli",
                 annotation = annotation,
                 legend_title = "Copy number \ngain")
```

It looks like we now have a prety good separation between tumor and normal cells based mostly on Chr8 gain. 
Before we finish, we'll make a UMAP showing where the final annotations are.

```{r}
# look at which cells are tumor/normal
new_classification_df |> 
  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

@@

```{r}
# export final TSV with annotations 
final_classifications <- classification_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    tumor_cell_classification = combined_cnv_classification
  ) |> 
  unique()
